Fast Sorting on Modern Computers

### 2005-07-31 docs/tech_report/radix.tex
fixing up lots of things
the grand reaming is almost complete

\subsection{Aligned Cache Radixsort}
In Section \ref{tiled mergesort}, mergesort is improved by aligning its array
with reference to it auxilliary array. Specifically, extra memory is allocated
for the auxilliary array, which is then offset so that the start of the two
arrays are exactly half the length of the cache apart. We further added an
alignment for the level 1 cache (see Section \ref{double tiled mergesort}) to
significantly reduce level 1 misses at the expense of a small number of level 2
misses. We adapt this strategy here, and call it Aligned Cache Radixsort.

\section{Results}
\subsection{Test parameters}
Keys are to be divided into segments 8 bits long, resulting in four sorting
steps to completely sort the array. The radix used is therefore $2^8 = 256$.

@@

Due to the reduced copying, cache radixsort should have a lower instruction
count and fewer cache misses that base radixsort. There should be nine passes
over the array, leading to just over one miss per key, as opposed to two misses
per key with base radixsort. The complexity and branch prediction results
should remain the same. The instruction count should, however, reduce by nearly
a half to $fract{9}{16}$ths of base radixsort.

Based on LaMarca's results, showing quicksort to be faster than radixsort by a
small amount, we expect cache radixsort to run faster than quicksort.

\subsection{Simulation results}
\plot{radix}{}{}{}{}{}{}

@@

Aligned cache radixsort performs exactly the same as cache radixsort in both
level 1 and level 2 misses, indicating that the alignment, and the associated
memory overhead, are not necessary.

@@

With the low number of cache misses, low instruction count and negligible branch
mispredictions, it is no surprise that cache radixsort performs as well as it
does. As shown in Figure \ref{radix cycles}, cache radixsorts performance is
almost linear, with only a slight increase when the arrays no longer fit in the
cache.  Even then, the increase in cache misses are more than made up for by the
lack of branch mispredictions, and quicksort ends up taking almost 50\% percent
longer to sort more than 4 million keys.

>>>>

\subsection{Aligned Memory-tuned Radixsort}
In Section \ref{tiled mergesort}, mergesort is improved by aligning its array
with reference to it auxilliary array. Specifically, extra memory is allocated
for the auxilliary array, which is then offset so that the start of the two
arrays are exactly half the length of the cache apart. We further added an
alignment for the level 1 cache (see Section \ref{double-aligned tiled
mergesort}) to significantly reduce level 1 misses at the expense of a small
number of level 2 misses. We adapt this strategy here, and call it Aligned Cache
Radixsort.

\section{Results}
\subsection{Test Parameters}
Keys are to be divided into segments 8 bits long, resulting in four sorting
steps to completely sort the array. The radix used is therefore $2^8 = 256$.

@@

Due to the reduced copying, memory-tuned radixsort should have a lower instruction
count and fewer cache misses that base radixsort. There should be nine passes
over the array, leading to just over one miss per key, as opposed to two misses
per key with base radixsort. The complexity and branch prediction results
should remain the same. The instruction count should, however, reduce by nearly
a half to $fract{9}{16}$ths of base radixsort.

Based on LaMarca's results, showing quicksort to be faster than radixsort by a
small amount, we expect memory-tuned radixsort to run faster than quicksort.

\subsection{Simulation Results}
\plot{radix}{}{}{}{}{}{}

@@

Aligned memory-tuned radixsort performs exactly the same as memory-tuned radixsort in both
level 1 and level 2 misses, indicating that the alignment, and the associated
memory overhead, are not necessary.

@@

With the low number of cache misses, low instruction count and negligible branch
mispredictions, it is no surprise that memory-tuned radixsort performs as well as it
does. As shown in Figure \ref{radix cycles}, memory-tuned radixsorts performance is
almost linear, with only a slight increase when the arrays no longer fit in the
cache.  Even then, the increase in cache misses are more than made up for by the
lack of branch mispredictions, and quicksort ends up taking almost 50\% percent
longer to sort more than 4 million keys.

### 2005-07-31 docs/tech_report/result_list.tex
the grand reaming is almost complete

\label{result_list}
\afterpage{
\smallplot{insertsort}{insertion sort}{}{}{}{}{}
\smallplot{selectsort}{selection sort}{}{}{}{}{}
\smallplot{bubblesort}{bubblesort}{}{}{}{}{}
\smallplot{bubblesort2}{improved bubblesort}{}{}{}{}{}
\smallplot{shakersort}{shakersort}{}{}{}{}{}
\smallplot{shakersort2}{improved shakersort}{}{}{}{}{}
%
\smallplot{base_heapsort}{base heapsort}{}{}{}{}{}
\smallplot{memory_tuned4_heapsort}{cache heapsort (4-heap)}{}{}{}{}{}
\smallplot{memory_tuned_heapsort}{cache heapsort (8-heap)}{}{}{}{}{}
%
\smallplot{algorithm_n}{algorithm n}{}{}{}{}{}
\smallplot{algorithm_s}{algorithm s}{}{}{}{}{}
\smallplot{base_mergesort}{base mergesort}{}{}{}{}{}
\smallplot{tiled_mergesort}{tiled mergesort}{}{}{}{}{}
\smallplot{multi_mergesort}{multi-mergesort}{}{}{}{}{}
\smallplot{double_aligned_mergesort}{double tiled mergesort}{}{}{}{}{}
\smallplot{double_aligned_multi_mergsort}{double multi-mergesort}{}{}{}{}{}
%
\smallplot{base_quicksort1}{base quicksort (no median)}{}{}{}{}{}
\smallplot{base_quicksort}{base quicksort (median-of-3)}{}{}{}{}{}
\smallplot{base_quicksort5}{base quicksort (pseudo-median-of-5)}{}{}{}{}{}
\smallplot{base_quicksort7}{base quicksort (pseudo-median-of-7)}{}{}{}{}{}
\smallplot{base_quicksort9}{base quicksort (pseudo-median-of-9)}{}{}{}{}{}
\smallplot{memory_tuned_quicksort}{memory-tuned quicksort}{}{}{}{}{}
\smallplot{multi_quicksort}{multi-quicksort (binary search)}{}{}{}{}{}
\smallplot{multi_quicksort_seq}{multi-quicksort (sequential search)}{}{}{}{}{}
%
\smallplot{base_radixsort}{base radixsort}{}{}{}{}{}
\smallplot{memory_tuned_radixsort}{cache radixsort}{}{}{}{}{}
\smallplot{aligned_memory_tuned_radixsort}{aligned cache radixsort}{}{}{}{}{}
%
\smallplot{shellsort}{shellsort}{}{}{}{}{}
\smallplot{improved_shellsort}{improved shellsort}{}{}{}{}{}
}

>>>>

\label{result_list}
\afterpage{
\smallplot{insertsort}{insertion sort}{}{}{}{}{}
\smallplot{selectsort}{selection sort}{}{}{}{}{}
\smallplot{bubblesort}{bubblesort}{}{}{}{}{}
\smallplot{improved_bubblesort}{improved bubblesort}{}{}{}{}{}
\smallplot{shakersort}{shakersort}{}{}{}{}{}
\smallplot{improved_shakersort}{improved shakersort}{}{}{}{}{}
%
\smallplot{base_heapsort}{base heapsort}{}{}{}{}{}
\smallplot{memory_tuned4_heapsort}{memory-tuned heapsort (4-heap)}{}{}{}{}{}
\smallplot{memory_tuned_heapsort}{memory-tuned heapsort (8-heap)}{}{}{}{}{}
%
\smallplot{algorithm_n}{algorithm n}{}{}{}{}{}
\smallplot{algorithm_s}{algorithm s}{}{}{}{}{}
\smallplot{base_mergesort}{base mergesort}{}{}{}{}{}
\smallplot{tiled_mergesort}{tiled mergesort}{}{}{}{}{}
\smallplot{multi_mergesort}{multi-mergesort}{}{}{}{}{}
\smallplot{double_aligned_mergesort}{double-aligned tiled mergesort}{}{}{}{}{}
\smallplot{double_aligned_multi_mergsort}{double-aligned multi-mergesort}{}{}{}{}{}
%
\smallplot{base_quicksort1}{base quicksort (no median)}{}{}{}{}{}
\smallplot{base_quicksort}{base quicksort (median-of-3)}{}{}{}{}{}
\smallplot{base_quicksort5}{base quicksort (pseudo-median-of-5)}{}{}{}{}{}
\smallplot{base_quicksort7}{base quicksort (pseudo-median-of-7)}{}{}{}{}{}
\smallplot{base_quicksort9}{base quicksort (pseudo-median-of-9)}{}{}{}{}{}
\smallplot{memory_tuned_quicksort}{memory-tuned quicksort}{}{}{}{}{}
\smallplot{multi_quicksort}{multi-quicksort (binary search)}{}{}{}{}{}
\smallplot{multi_quicksort_seq}{multi-quicksort (sequential search)}{}{}{}{}{}
%
\smallplot{base_radixsort}{base radixsort}{}{}{}{}{}
\smallplot{memory_tuned_radixsort}{memory-tuned radixsort}{}{}{}{}{}
\smallplot{aligned_memory_tuned_radixsort}{aligned memory-tuned radixsort}{}{}{}{}{}
%
\smallplot{shellsort}{shellsort}{}{}{}{}{}
\smallplot{improved_shellsort}{improved shellsort}{}{}{}{}{}
}

### 2005-07-31 docs/tech_report/shell.tex
fixing up lots of things
the grand reaming is almost complete

\section{Results}
\subsection{Test parameters}
The increments in Gonnet's shellsort are $\frac{5}{11}N$, $(\frac{5}{11})^2N$,
$(\frac{5}{11})^3N$ ... $1$;

@@

\subsection{Simulation results}
\plot{shell}{}{}{}{}{}{}

@@

Shellsort's cache misses are certainly in the range of an $O(NlogN)$ sort, and
are not considerably worse than cache heapsort, and are in fact better than base
heapsort. This occurs without any cache improvements being performed upon them.
Shellsort's branch mispredictions, are, as predicted $O(Nlog_{\frac{5}{11}}N)$.

@@

%TODO
\section{Predictor results}

>>>>

\section{Results}
\subsection{Test Parameters}
The increments in Gonnet's shellsort are $\frac{5}{11}N$, $(\frac{5}{11})^2N$,
$(\frac{5}{11})^3N$ ... $1$;

@@

\subsection{Simulation Results}
\plot{shell}{}{}{}{}{}{}

@@

Shellsort's cache misses are certainly in the range of an $O(NlogN)$ sort, and
are not considerably worse than memory-tuned heapsort, and are in fact better
than base heapsort. This occurs without any cache improvements being performed
upon them.  Shellsort's branch mispredictions, are, as predicted
$O(Nlog_{\frac{5}{11}}N)$.

@@

%TODO
\section{Predictor Results}

### 2005-07-31 docs/tech_report/software_predictor.tex
the grand reaming is almost complete

\afterpage{
\sixplot{base_heapsort_0.eps}{memory_tuned4_heapsort_0.eps}{memory_tuned4_heapsort_1.eps}{memory_tuned_heapsort_0.eps}{memory_tuned_heapsort_1.eps}{}{base and cache heapsorts}

\sixplot{base_quicksort_0.eps}{base_quicksort9_0.eps}{memory_tuned_quicksort_0.eps}{multi_quicksort_0.eps}{sequential_multi_quicksort_0.eps}{}{base, cache and multi quicksorts and base quicksort9}
\sixplot{base_quicksort1_0.eps}{base_quicksort_0.eps}{base_quicksort5_0.eps}{base_quicksort7_0.eps}{base_quicksort9_0.eps}{}{base quicksort without a median, with median of 3, and with pseudo median of 5, 7 and 9}

\sixplot{algorithm_n_0.eps}{algorithm_n_1.eps}{algorithm_s_0.eps}{algorithm_s_1.eps}{}{}{Knuth's Mergesorts}
\sixplot{base_mergesort_0.eps}{base_mergesort_1.eps}{base_mergesort_2.eps}{base_mergesort_3.eps}{base_mergesort_4.eps}{base_mergesort_5.eps}{base mergesort}
\sixplot{double_aligned_mergesort_0.eps}{double_aligned_mergesort_1.eps}{double_aligned_mergesort_2.eps}{double_aligned_mergesort_3.eps}{double_aligned_mergesort_4.eps}{double_aligned_mergesort_5.eps}{double tiled mergesort - part 1}
\sixplot{double_aligned_mergesort_6.eps}{double_aligned_mergesort_7.eps}{double_aligned_multi_mergsort_0.eps}{double_aligned_multi_mergsort_1.eps}{double_aligned_multi_mergsort_2.eps}{double_aligned_multi_mergsort_3.eps}{double tiled mergesort - part 2, and double multi mergesort - part 1}
\sixplot{double_aligned_multi_mergsort_4.eps}{double_aligned_multi_mergsort_5.eps}{double_aligned_multi_mergsort_6.eps}{double_aligned_multi_mergsort_7.eps}{double_aligned_multi_mergsort_8.eps}{double_aligned_multi_mergsort_9.eps}{double multi-mergesort - part 2}

>>>>

\afterpage{
\sixplot{base_heapsort_0.eps}{memory_tuned4_heapsort_0.eps}{memory_tuned4_heapsort_1.eps}{memory_tuned_heapsort_0.eps}{memory_tuned_heapsort_1.eps}{}{base and memory-tuned heapsorts}

\sixplot{base_quicksort_0.eps}{base_quicksort9_0.eps}{memory_tuned_quicksort_0.eps}{multi_quicksort_0.eps}{sequential_multi_quicksort_0.eps}{}{base, memory-tuned and multi quicksorts and base quicksort (pseudo-median-of-9)}
\sixplot{base_quicksort1_0.eps}{base_quicksort_0.eps}{base_quicksort5_0.eps}{base_quicksort7_0.eps}{base_quicksort9_0.eps}{}{base quicksort without a median, with median of 3, and with pseudo-medians of 5, 7 and 9}
\sixplot{algorithm_n_0.eps}{algorithm_n_1.eps}{algorithm_s_0.eps}{algorithm_s_1.eps}{}{}{Knuth's mergesorts}
\sixplot{base_mergesort_0.eps}{base_mergesort_1.eps}{base_mergesort_2.eps}{base_mergesort_3.eps}{base_mergesort_4.eps}{base_mergesort_5.eps}{base mergesort}
\sixplot{double_aligned_mergesort_0.eps}{double_aligned_mergesort_1.eps}{double_aligned_mergesort_2.eps}{double_aligned_mergesort_3.eps}{double_aligned_mergesort_4.eps}{double_aligned_mergesort_5.eps}{double-aligned tiled mergesort - part 1}
\sixplot{double_aligned_mergesort_6.eps}{double_aligned_mergesort_7.eps}{double_aligned_multi_mergsort_0.eps}{double_aligned_multi_mergsort_1.eps}{double_aligned_multi_mergsort_2.eps}{double_aligned_multi_mergsort_3.eps}{double-aligned tiled mergesort - part 2, and double multi mergesort - part 1}
\sixplot{double_aligned_multi_mergsort_4.eps}{double_aligned_multi_mergsort_5.eps}{double_aligned_multi_mergsort_6.eps}{double_aligned_multi_mergsort_7.eps}{double_aligned_multi_mergsort_8.eps}{double_aligned_multi_mergsort_9.eps}{double-aligned multi-mergesort - part 2}
